Bayesian Statistical Modeling with Predictors from LLMs

### 2024-03-30 04-paper/paper-LLM-probmod.tex
completed §6 and revised appendix C

\mf{UP TO HERE}

@@

All previous results where obtained for LMM predictions derived from GPT-3.5 turbo.
To investigate whether key results replicate for other LLMs and scales, we ran the exact same analyses for item- and condition-level data also for scores derived from variants of LLaMA2 \citep{TouvronLavril2023:LLaMA:-Open-and}, in particular the 7, 13 and 70 billion parameter versions of the versions fine-tuned with human feedback only (marked as `hf' in the following) and based on human feedback and chat data (marked as `chat').
Table~\ref{tab:sumStats-LLaMA2} shows the relevant summary statistics and Bayesian posterior predictive $p$-values.
Figure~\ref{fig:LLaMA-results-vPPC} shows visual posterior predictive checks for the condition-level data.
The plots show the differences between human counts for each condition and option and the models' posterior predictions.
We find that the average-WTA predictor is consistently able to capture the data from the production condition except for one of the six models (LLaMA2-chat 70B).
However, the average-WTA predictor only recovers the interpretation data for only one of the six models (LLaMA2-chat 13B).
Interestingly, for this LLM as backend for scoring, also the average-probability predictor is able to recover the interpretation data, unlike for the GPT-3.5 backend.

\mf{TODO: item-level; check estimated alphas for condition-level!?}

In sum, we find general support for the previous conclusion that the average-WTA predictor is most successful in capturing the condition-level data, but we also find variability in which method of aggregation works well with which LLM as scoring model.

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Average-probability vs.~average-WTA for accuracy scores}
\label{sec:aver-prob-vs}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Standard benchmark testing looks at \emph{task accuracy}, defined as the probability of selecting the ``gold standard'' response, usually based on a WTA selection of the highest scoring option.
We can generalize this and define the \emph{softmax-based accuracy}, $P_{\text{cond}}^{\text{SM}}\left(r_{1} \right)$, as the average probability of choosing the designated target response option $r_{1}$ for the \emph{condition-level soft-max prediction}:
\begin{align*}
  P_{\text{cond}}^{\text{SM}}\left(r_{i} \right) = \frac{1}{m} \sum_{k = 1}^{m} P_{\text{item}}^{\text{SM}} \left(y_{ki} \right)\,.
\end{align*}
As before, the average-WTA task accuracy is the special case of $\alpha \rightarrow\infty$.

Indeed, WTA-based accuracy can differ qualitatively from softmax-based accuracy, as shown by the following example.
The example shows that differences between accuracy measures depend on the variation in item-level scores, in particular on the relation between score-ordering and score-differences.
Note that the example holds equally if numbers for the two options are reversed, so that there is no way of saying which of the two measures of accuracy would generally be more favorable for selecting the target option.
%
\begin{quote}
  % \ex \label{bsp:exmpl-diff-WTA-SM}
  \textbf{Example:}
  Imagine that there are two options, and that the target option's score is a small $\epsilon$ higher in 80\% of the task's items, and otherwise lower.
  The WTA-based accuracy is 0.8.
  This number is useful as a performance measure for applications in which the LLM is used in exactly the way the WTA strategy describes, e.g., any implementation which is outcome equivalent to greedy decoding with rejection sampling on a domain that contains only the available options.
  For such a case, it never matters how much worse the goal answer is scored in the 20\% of the cases where it is not the maximum.
  As only the best option will be chosen, that information is irrelevant.
  But if an application uses anything other than greedy-like responses, the accuracy score of 0.8 may be misleading.
  If the remaining 20\% of the items are such that the non-goal option is almost infinitely better, it would be chosen under a pure sampling strategy, where $\alpha = 1$, with virtual certainty, so the softmax-based accuracy would around 0.4.\footnote{The probability of the target option in the 80\% of items where the goal answer is slightly better is 0.5 in the limit of $\epsilon \rightarrow 0$, and it is virtually 0 in the remaining 20\% of the cases. This gives an expected rate of: $\nicefrac{4}{5} \ \nicefrac{1}{2} + \nicefrac{1}{5} \ 0 = \nicefrac{2}{5}$.}
\end{quote}
%
The upshot of these considerations is that the standard practice of WTA-based performance assessment for LLMs gives false, or at least misleading or inaccurate results, whenever not all downstream applications use a greedy-like sampling strategy (which is almost certainly the case), and there is variability in item-level predictions.

>>>>

@@

@@

All previous results where obtained for LMM predictions derived from GPT-3.5 turbo.
To investigate whether key results replicate for other LLMs and scales, we ran the same analyses also for scores derived from variants of LLaMA2 \citep{TouvronLavril2023:LLaMA:-Open-and}, in particular the 7, 13 and 70 billion parameter versions of the models fine-tuned with human feedback only (marked as `hf' in the following) and based on human feedback and chat data (marked as `chat').
Table~\ref{tab:sumStats-LLaMA2} shows the relevant summary statistics and Bayesian posterior predictive $p$-values.
Figure~\ref{fig:LLaMA-results-vPPC} shows visual posterior predictive checks for the condition-level data.
The plots show the \emph{differences} between observed counts and the models' posterior predictions for expected counts for each condition and response option.
We find that the average-WTA predictor is consistently able to capture the data from the production condition, except for one of the six models (LLaMA2-chat 70B).
However, the average-WTA predictor recovers the interpretation data for only one of the six models (LLaMA2-chat 13B, see also Table~\ref{tab:sumStats-LLaMA2}).
Interestingly, for LLaMA2-chat 13B also the average-probability predictor is able to recover the interpretation data, unlike for the GPT-3.5.
As for GPT-3.5, no model captured the item-level variance observed in the human data.
In sum, we find general support for the previous conclusions that the average-WTA predictor is most successful in capturing the condition-level data, and therefore that, where models manage to recover the condition-level data, they seem to be ``right for the wrong reasons.''
Additionally, we also find variability in which method of aggregation works well with which LLM as scoring model.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Excursion: Accuracy scores from average-probability vs.~average-WTA predictors}
\label{sec:aver-prob-vs}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Standard benchmark testing looks at \emph{task accuracy}, defined as the probability of selecting the ``gold standard'' response, usually based on ``winner-takes-all'' (WTA) selection of the highest scoring option.
We can generalize this and define the \emph{softmax-based accuracy} as the average probability of choosing the designated target response option $r_{1}$ for the \emph{condition-level soft-max prediction} (notation as defined in main text):
\begin{align*}
  P_{\text{cond}}^{\text{SM}}\left(r_{1} \right) = \frac{1}{m} \sum_{k = 1}^{m} P_{\text{item}}^{\text{SM}} \left(y_{k1} \right)\,.
\end{align*}
The WTA-based accuracy (the standard measure in most benchmark testing) is the special case of $\alpha \rightarrow\infty$.

The WTA-based accuracy can differ qualitatively from the more general softmax-based accuracy, as shown by the following example.
The example shows that differences between accuracy measures depend on the variation in item-level scores, in particular on the relation between score-ordering and score-differences.
Note that the example holds equally if numbers for the two options are reversed, so that there is no way of saying which of the two measures of accuracy would generally be more favorable for selecting the target option.
%
\begin{quote}
  % \ex \label{bsp:exmpl-diff-WTA-SM}
  \textbf{Example:}
  Imagine that there are two options, and that the target option's score is a small $\epsilon$ higher in 80\% of the task's items, and otherwise lower.
  The WTA-based accuracy is 0.8.
  This number is useful as a performance measure for applications in which the LLM is used in exactly the way the WTA strategy describes, e.g., any implementation which is outcome equivalent to greedy decoding with rejection sampling on a domain that contains only the available options.
  For such a case, it never matters how much worse the goal answer is scored in the 20\% of the cases where it is not the maximum.
  As only the best option will be chosen, that information is irrelevant.
  But if an application uses anything other than greedy-like responses, the accuracy score of 0.8 may be misleading.
  If the remaining 20\% of the items are such that the non-goal option is almost infinitely better, it would be chosen under a pure sampling strategy, where $\alpha = 1$, with virtual certainty, so the softmax-based accuracy would around 0.4.\footnote{The probability of the target option in the 80\% of items where the goal answer is slightly better is 0.5 in the limit of $\epsilon \rightarrow 0$, and it is virtually 0 in the remaining 20\% of the cases. This gives an expected rate of: $\nicefrac{4}{5} \ \nicefrac{1}{2} + \nicefrac{1}{5} \ 0 = \nicefrac{2}{5}$.}
\end{quote}
%
The upshot of these considerations is that the standard practice of WTA-based performance assessment for LLMs gives false, or at least misleading or inaccurate results, whenever not all downstream applications use a greedy-like sampling strategy (which is almost certainly the case), and there is variability in item-level predictions.
